---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

There are two ways to group in dplyr:

-   Persistent grouping with [group_by()]

-   Per-operation grouping with `.by`/`by`

This help page is dedicated to explaining where and why you might want to use the latter.

Depending on the dplyr verb, the per-operation grouping argument may be named `.by` or `by`.
The *Supported verbs* section below outlines this on a case-by-case basis.
The remainder of this page will refer to `.by` for simplicity.

Grouping radically affects the computation of the dplyr verb you use it with, and one of the goals of `.by` is to allow you to place that grouping specification alongside the code that actually uses it.
As an added benefit, with `.by` you no longer need to remember to [ungroup()] after [summarise()], and `summarise()` won't ever message you about how it's handling the groups!

This idea comes from [data.table](https://CRAN.R-project.org/package=data.table), which allows you to specify `by` alongside modifications in `j`, like: `dt[, .(x = mean(x)), by = g]`.

### Supported verbs

-   [`mutate(.by = )`][mutate()]

-   [`summarise(.by = )`][summarise()]

-   [`reframe(.by = )`][reframe()]

-   [`filter(.by = )`][filter()]

-   [`slice(.by = )`][slice()]

-   [`slice_head(by = )`][slice_head()] and [`slice_tail(by = )`][slice_tail()]

-   [`slice_min(by = )`][slice_min()] and [`slice_max(by = )`][slice_max()]

-   [`slice_sample(by = )`][slice_sample()]

Note that some dplyr verbs use `by` while others use `.by`.
This is a purely technical difference.

### Differences between `.by` and `group_by()`

| `.by`                                                   | `group_by()`                                                       |
|---------------------------------------------------------|--------------------------------------------------------------------|
| Grouping only affects a single verb                     | Grouping is persistent across multiple verbs                       |
| Selects variables with [tidy-select][dplyr_tidy_select] | Computes expressions with [data-masking][rlang::args_data_masking] |
| Summaries use existing order of group keys              | Summaries sort group keys in ascending order                       |

### Using `.by`

Let's take a look at the two grouping approaches using this `expenses` data set, which tracks costs accumulated across various `id`s and `region`s:

```{r}
expenses <- tibble(
  id = c(1, 2, 1, 3, 1, 2, 3),
  region = c("A", "A", "A", "B", "B", "A", "A"),
  cost = c(25, 20, 19, 12, 9, 6, 6)
)
expenses
```

Imagine that you wanted to compute the average cost per region.
You'd probably write something like this:

```{r}
expenses %>%
  group_by(region) %>%
  summarise(cost = mean(cost))
```

Instead, you can now specify the grouping *inline* within the verb:

```{r}
expenses %>%
  summarise(cost = mean(cost), .by = region)
```

`.by` applies to a single operation, meaning that since `expenses` was an ungrouped data frame, the result after applying `.by` will also always be an ungrouped data frame, regardless of the number of grouping columns.

```{r}
expenses %>%
  summarise(cost = mean(cost), .by = c(id, region))
```

Compare that with `group_by() %>% summarise()`, where `summarise()` generally peels off 1 layer of grouping by default, typically with a message that it is doing so:

```{r}
expenses %>%
  group_by(id, region) %>%
  summarise(cost = mean(cost))
```

Because `.by` grouping applies to a single operation, you don't need to worry about ungrouping, and it never needs to emit a message to remind you what it is doing with the groups.

Note that with `.by` we specified multiple columns to group by using the [tidy-select][dplyr_tidy_select] syntax `c(id, region)`.
If you have a character vector of column names you'd like to group by, you can do so with `.by = all_of(my_cols)`.
It will group by the columns in the order they were provided.

To prevent surprising results, you can't use `.by` on an existing grouped data frame:

```{r, error=TRUE}
expenses %>% 
  group_by(id) %>%
  summarise(cost = mean(cost), .by = c(id, region))
```

So far we've focused on the usage of `.by` with `summarise()`, but `.by` works with a number of other dplyr verbs.
For example, you could append the mean cost per region onto the original data frame as a new column rather than computing a summary:

```{r}
expenses %>%
  mutate(cost_by_region = mean(cost), .by = region)
```

Or you could slice out the maximum cost per combination of id and region:

```{r}
# Note that the argument is named `by` in `slice_max()`
expenses %>%
  slice_max(cost, n = 1, by = c(id, region))
```

### Result ordering

When used with `.by`, `summarise()`, `reframe()`, and `slice()` all maintain the ordering of the existing data.
This is different from `group_by()`, which has always sorted the group keys in ascending order.

```{r}
df <- tibble(
  month = c("jan", "jan", "feb", "feb", "mar"),
  temp = c(20, 25, 18, 20, 40)
)

# Uses ordering by "first appearance" in the original data
df %>%
  summarise(average_temp = mean(temp), .by = month)

# Sorts in ascending order
df %>%
  group_by(month) %>%
  summarise(average_temp = mean(temp))
```

If you need sorted group keys, we recommend that you explicitly use [arrange()] either before or after the call to `summarise()`, `reframe()`, or `slice()`.
This also gives you full access to all of `arrange()`'s features, such as `desc()` and the `.locale` argument.

### Verbs without `.by` support

If a dplyr verb doesn't support `.by`, then that typically means that the verb isn't inherently affected by grouping.
For example, [pull()] and [rename()] don't support `.by`, because specifying columns to group by would not affect their implementations.

That said, there are a few exceptions to this where sometimes a dplyr verb doesn't support `.by`, but *does* have special support for grouped data frames created by [group_by()].
This is typically because the verbs are required to retain the grouping columns, for example:

-   [select()] always retains grouping columns, with a message if any aren't specified in the `select()` call.

-   [distinct()] and [count()] place unspecified grouping columns at the front of the data frame before computing their results.

-   [arrange()] has a `.by_group` argument to optionally order by grouping columns first.

If `group_by()` didn't exist, then these verbs would not have special support for grouped data frames.
