<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Baseball benchmarks}
-->

```{r, echo = FALSE, message = FALSE}
library(dplyr)
library(microbenchmark)
library(data.table)
library(Lahman)
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE
)

options(digits = 3)
```

# Benchmarks: baseball data

The purpose of these benchmarks is to be as fair as possible, to help understand the relatively performance tradeoffs of the different approaches. If you think my implementation of base or data.table equivalents is suboptimal, please let me know better ways.

Also note that I consider any significant performance difference between `dt` and `dt_raw` to be a bug in dplyr: for individual operations there should be very little overhead to calling data.table via dplyr. However, data.table may be significantly faster when performing the same sequence of operations as dplyr. This is because currently dplyr uses an eager evaluation approach so the individual calls to `[.data.table` don't get as much information about the desired result as the single call to `[.data.table` would if you did it by hand.

## Data setup

The following benchmarks explore the performance on a somewhat realistic example: the `Batting` dataset from the Lahman package. It contains `r nrow(Batting)` records on the batting careers of `r length(Batting$playerID)` players from `r min(Batting$yearID)` to `r max(Batting$yearID)`.

The first code block defines three alternative backends for the Batting dataset, and a players dataset that represents operations to be performed by player:

```{r setup}
batting_df <- tbl_df(Batting)
players_df <- group_by(batting_df, playerID)

batting_dt <- tbl_dt(Batting)
players_dt <- group_by(batting_dt, playerID)
```

## Summarise

Compute the average number of at bats for each player:

```{r summarise-mean}
microbenchmark(
  dplyr_df = summarise(players_df, ab = mean(AB)),
  dplyr_dt = summarise(players_dt, ab = mean(AB)),
  dt_raw =   players_dt[, list(ab = mean(AB)), by = playerID],
  base =     tapply(batting_df$AB, batting_df$playerID, FUN = mean),
  times = 5, 
  unit = "ms"
)
```

NB: base implementation captures computation but not output format, giving considerably less output.

However, this comparison is slightly unfair because both data.table and `summarise.tbl_df` use tricks to find a more efficient implementation of `mean()`. Data table calls a `C` implementation of the `mean (using `.External(Cfastmean, B, FALSE)`  and thus avoiding the overhead of S3 method dispatch). `dplyr::summarise` uses a hybrid evaluation technique, where common functions are implemented purely in C++, avoiding R function call overhead.

```{r sumarise-mean_}
mean_ <- function(x) .Internal(mean(x))
microbenchmark(
  dplyr_df = summarise(players_df, ab = mean_(AB)),
  dplyr_dt = summarise(players_dt, ab = mean_(AB)),
  dt_raw =   players_dt[, list(ab = mean_(AB)), by = playerID],
  base =     tapply(batting_df$AB, batting_df$playerID, FUN = mean_),
  times = 5, 
  unit = "ms"
)
```

## Arrange

Arrange by year within each player:

```{r arrange}
microbenchmark(
  dplyr_df = arrange(players_df, yearID),
  dplyr_dt = arrange(players_dt, yearID),
  dt_raw =   batting_dt[order(playerID, yearID), ],
  base   =   batting_df[order(batting_df$playerID, batting_df$yearID), ],
  times = 2, 
  unit = "ms"
)  
```

## Filter

Find the year for which each player played the most games:

```{r filter}
microbenchmark(
  dplyr_df = filter(players_df, G == max(G)),
  dplyr_dt = filter(players_dt, G == max(G)),
  base   =   batting_df[ave(batting_df$G, batting_df$playerID, FUN = max) ==
    batting_df$G, ],
  times = 2, 
  unit = "ms"  
)
```

I'm not aware of a single line data table equivalent ([see SO 16573995](http://stackoverflow.com/questions/16573995/)). Suggetions welcome.  dplyr currently doesn't support hybrid evaluation for logical comparison, but it is scheduled for 0.2 (see [#113](https://github.com/hadley/dplyr/issues/113)), this should give an additional (10-20x) speed up.

## Mutate

Rank years based on number of at bats:

```{r mutate}
microbenchmark(
  dplyr_df  = mutate(players_df, rank = rank(desc(AB))),
  dplyr_dt  = mutate(players_dt, rank = rank(desc(AB))),
  dt_raw =    players_dt[, list(rank = rank(desc(AB))), by = playerID],
  times = 2, 
  unit = "ms"
)
```

Compute year of career:

```{r mutate2}
microbenchmark(
  dplyr_df = mutate(players_df, cyear = yearID - min(yearID) + 1),
  dplyr_dt = mutate(players_dt, cyear = yearID - min(yearID) + 1),
  dt_raw =   players_dt[, list(cyear = yearID - min(yearID) + 1), by = playerID],
  times = 5, 
  unit = "ms"
)
```

Rank is a relatively expensive operation and `min()` is relatively cheap, showing the the relative performance overhead of the difference techniques.

dplyr currently has some support for hybrid evaluation of window functions:

```{r mutate_hybrid}
microbenchmark(
  dplyr_df  = mutate(players_df, rank = min_rank(AB)),
  dplyr_dt  = mutate(players_dt, rank = min_rank(AB)),
  dt_raw =    players_dt[, list(rank = min_rank(AB)), by = playerID],
  times = 2, 
  unit = "ms"
)
```

## Joins

We conclude with some quick comparisons of joins. First we create two new datasets: `master` which contains demographic information on each player, and `hall_of_fame` which contains all players inducted into the hall of fame.

```{r}
master_df <- tbl_df(Master) %.% select(playerID, hofID, birthYear)
hall_of_fame_df <- tbl_df(HallOfFame) %.% filter(inducted == "Y") %.% 
  select(hofID, votedBy, category)

master_dt <- tbl_dt(Master) %.% select(playerID, hofID, birthYear)
hall_of_fame_dt <- tbl_dt(HallOfFame) %.% filter(inducted == "Y") %.% 
  select(hofID, votedBy, category)
```


```{r}
microbenchmark(
  dplyr_df = left_join(master_df, hall_of_fame_df, by = "hofID"),
  dplyr_dt = left_join(master_dt, hall_of_fame_dt, by = "hofID"),
  base     = merge(master_df, hall_of_fame_df, by = "hofID", all.x = TRUE),
  times = 10, 
  unit = "ms"
)

microbenchmark(
  dplyr_df = inner_join(master_df, hall_of_fame_df, by = "hofID"),
  dplyr_dt = inner_join(master_dt, hall_of_fame_dt, by = "hofID"),
  base     = merge(master_df, hall_of_fame_df, by = "hofID"),
  times = 10, 
  unit = "ms"
)

microbenchmark(
  dplyr_df = semi_join(master_df, hall_of_fame_df, by = "hofID"),
  dplyr_dt = semi_join(master_dt, hall_of_fame_dt, by = "hofID"),
  times = 10, 
  unit = "ms"
)

microbenchmark(
  dplyr_df = anti_join(master_df, hall_of_fame_df, by = "hofID"),
  dplyr_dt = anti_join(master_dt, hall_of_fame_dt, by = "hofID"),
  times = 10, 
  unit = "ms"
)
```
