---
title: "Data frames"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(dplyr.print_min = 4L, dplyr.print_max = 4L)
library(dplyr)
```

## Creating

`data_frame()` is a nice way to create data frames. It encapsulates best practices for data frames:

  * It never changes the type of its inputs (i.e. no more `stringsAsFactors = FALSE`!)
    
    ```{r}
    data.frame(x = letters) %>% sapply(class)
    data_frame(x = letters) %>% sapply(class)
    ```
    
    This makes it easier to use with list-columns:
    
    ```{r}
    data_frame(x = 1:3, y = list(1:5, 1:10, 1:20))
    ```
    
    List-columns are most commonly created by `do()`, but they can be useful to
    create by hand.
      
  * It never adjusts the names of variables:
  
    ```{r}
    data.frame(`crazy name` = 1) %>% names()
    data_frame(`crazy name` = 1) %>% names()
    ```

  * It evaluates its arguments lazyily and in order:
  
    ```{r}
    data_frame(x = 1:5, y = x ^ 2)
    ```

  * It adds `tbl_df()` class to output so that if you accidentaly print a large 
    data frames you only get the first few rows.
    
    ```{r}
    data_frame(x = 1:5) %>% class()
    ```
    
  * It never uses `row.names()`, because the whole point of tidy data is to 
    store variables in a consistent way, so we shouldn't put one variable in a 
    special attribute.
  
  * It only recycles vectors of length 1. Recycling vectors of other lengths 
    is a frequent source of bugs in my experience.

## Coercion

To complement `data_frame()`, dplyr provides `as_data_frame()` for coercing lists into data frames. It does two things:

* Checks that the input list is valid for a data frame, i.e. that each element
  is named and of the equal length.
  
* Sets the class and attributes of the list to make it behave like a data frame.
  This modification does not require a deep copy of the input list, so is
  very fast.
  
This is much simpler than `as.data.frame()`. It's hard to explain precisely what `as.data.frame()` does, but similar to `do.call(cbind, lapply(x, data.frame))` - i.e. it coerces each component to a data frame and then `cbinds()` them all together. Consequently `as_data_frame()` is much faster than `as.data.frame()`:

```{r}
l2 <- replicate(26, sample(100), simplify = FALSE)
names(l2) <- letters
microbenchmark::microbenchmark(
  as_data_frame(l2),
  as.data.frame(l2)
)
```
