---
title: "Grouped data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grouped data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
```

dplyr verbs are particularly powerful when you apply them to grouped data frames (`grouped_df` objects). This vignette shows you:

* How to group, inspect, and ungroup with `group_by()` and friends.
  
* How individual dplyr verbs changes their behaviour when applied to grouped 
  data frame.

* How to access data about the "current" group from within a verb.

We'll start by loading dplyr:

```{r, message = FALSE}
library(dplyr)
```

## `group_by()`

The most important grouping verb is `group_by()`: it takes a data frame and one or more variables to group by:

```{r}
by_species <- starwars %>% group_by(species)
by_sex_gender <- starwars %>% group_by(sex, gender)
```

You can see the grouping when you print the data:

```{r}
by_species
by_sex_gender
```

Or use `tally()` to count the number of rows in each group. The `sort` argument is useful if you want to see the largest groups up front.

```{r}
by_species %>% tally()

by_sex_gender %>% tally(sort = TRUE)
```
As well as grouping by existing variables, you can group by any function of existing variables. This is equivalent to performing a `mutate()` **before** the `group_by()`:  

```{r group_by_with_expression}
bmi_breaks <- c(0, 18.5, 25, 30, Inf)

starwars %>%
  group_by(bmi_cat = cut(mass/(height/100)^2, breaks=bmi_breaks)) %>%
  tally()
```

## Group metadata

You can see underlying group data with `group_keys()`. It has one row for each group and one column for each grouping variable:

```{r group_vars}
by_species %>% group_keys()

by_sex_gender %>% group_keys()
```

You can see which group each row belongs to with `group_indices()`:

```{r}
by_species %>% group_indices()
```

And which rows each group contains with `group_rows()`:

```{r}
by_species %>% group_rows() %>% head()
```
Use `group_vars()` if you just want the names of the grouping variables:

```{r}
by_species %>% group_vars()
by_sex_gender %>% group_vars()
```

### Changing and adding to grouping variables

If you apply `group_by()` to an already grouped dataset, will overwrite the existing grouping variables. For example, the following code groups by `homeworld` instead of `species`:

```{r}
by_species %>%
  group_by(homeworld) %>%
  tally()
```

To **augment** the grouping, using `.add = TRUE`[^add]. For example, the following code groups by species and homeworld:

```{r}
by_species %>%
  group_by(homeworld, .add = TRUE) %>%
  tally()
```

[^add]: Note that the argument changed from `add = TRUE` to `.add = TRUE` in dplyr 1.0.0.

### Removing grouping variables

To remove all grouping variables, use `ungroup()`:

```{r}
by_species %>%
  ungroup() %>%
  tally()
```

You can also choose to selectively ungroup by listing the variables you want to remove:

```{r}
by_sex_gender %>% 
  ungroup(sex) %>% 
  tally()
```

## Verbs

The following sections describe how grouping affects the main dplyr verbs.

### `summarise()`

`summarise()` computes a summary for each group. This means that it starts from `group_keys()`, adding summary variables to the right hand side:

```{r summarise}
by_species %>%
  summarise(
    n = n(),
    height = mean(height, na.rm = TRUE)
  )
```

For historical reasons, `summarise()` removes the grouping variable on the right hand side:

```{r}
by_sex_gender %>% 
  summarise(n = n()) %>% 
  group_vars()
```

If there's only one grouping variable, the output will be ungrouped (i.e. a regular tibble).

### `select()`, `rename()`, and `relocate()`

`rename()` and `relocate()` behave identically with grouped and ungrouped data because they only affect the name or position of existing columns. Grouped `select()` is almost identical to ungrouped select, except that it always includes the grouping variables:

```{r select}
by_species %>% select(mass)
```

If you don't want the grouping variables, you'll have to first `ungroup()`. (This design is possibly a mistake, but we're stuck with it for now.)

### `arrange()`

Grouped `arrange()` is the same as ungrouped `arrange()`, unless you set `.by_group = TRUE`,
in which case it will order first by the grouping variables.

```{r}
by_species %>%
  arrange(desc(mass)) %>%
  relocate(species, mass)

by_species %>%
  arrange(desc(mass), .by_group = TRUE) %>%
  relocate(species, mass)
```

Note that second example is sorted by `species` (from the `group_by()` statement) and
then by `mass` (within species).

### `mutate()` and `transmute()`

In simple cases with vectorised functions, grouped and ungrouped `mutate()` give the same results. They differ when used with summary functions:

```{r by_homeworld}
# Subtract off global mean
starwars %>% 
  select(name, homeworld, mass) %>% 
  mutate(standard_mass = mass - mean(mass, na.rm = TRUE))

# Subtract off homeworld mean
starwars %>% 
  select(name, homeworld, mass) %>% 
  group_by(homeworld) %>% 
  mutate(standard_mass = mass - mean(mass, na.rm = TRUE))
```

Or with window functions like `min_rank()`:

```{r}
# Overall rank
starwars %>% 
  select(name, homeworld, height) %>% 
  mutate(rank = min_rank(height))

# Rank per homeworld
starwars %>% 
  select(name, homeworld, height) %>% 
  group_by(homeworld) %>% 
  mutate(rank = min_rank(height))
```

### `filter()`

A grouped `filter()` effectively does a `mutate()` to generate a logical variable, and then only keeps the rows where the variable is `TRUE`. This means that grouped filters can be used with summary functions. For example, we can find the tallest character of each species:

```{r filter}
by_species %>%
  select(name, species, height) %>% 
  filter(height == max(height))
```

You can also use `filter()` to remove entire groups. For example, the following code eliminates all groups that only have a single member:

```{r filter_group}
by_species %>%
  filter(n() != 1) %>% 
  tally()
```

### `slice()` and friends

`slice()` and friends (`slice_head()`, `slice_tail()`, `slice_sample()`, `slice_min()` and `slice_max()`) select rows within a group. For example, we can select the first observation within each species:

```{r slice}
by_species %>%
  relocate(species) %>% 
  slice(1)
```

Similarly, we can use `slice_min()` to select the smallest `n` values of a variable:

```{r slice_min}
by_species %>%
  filter(!is.na(height)) %>% 
  slice_min(height, n = 2)
```

## Computing on grouping information

Inside dplyr verbs, you can access various properties of the "current" group using a family of functions with the `cur_` prefix. These functions are typically needed for everyday usage of dplyr, but can be useful because they allow you to free from some of the typical constraints of dplyr verbs.

### `cur_data()`

`cur_data()` returns the current group, excluding grouping variables. It's useful to feed to functions that take a whole data frame. For example, the following code fits a linear model of `mass ~ height` to each species:

```{r cur_data}
by_species %>%
  filter(n() > 1) %>% 
  mutate(mod = list(lm(mass ~ height, data = cur_data())))
```

### `cur_group()` and `cur_group_id()`

`cur_group_id()` gives a unique numeric identifier for the current group. This is sometimes useful if you want to index into an external data structure.

```{r cur_group_id}
by_species %>%
  arrange(species) %>% 
  select(name, species, homeworld) %>% 
  mutate(id = cur_group_id())
```

