<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Adding new database support to dplyr}
-->

# Adding a new sql backend

This document describes how to describe a new SQL backend to dplyr. It's a work in progress, but will hopefully get started. If you get stuck or need more help, please email the [mailing list](mailto:manipulatr@googlegroups.com) and I'll help you get unstuck.

dplyr uses the S3 object system, so make sure that you're [familiar with the basics](http://adv-r.had.co.nz/OO-essentials.html#s3) before you start.

## Create the src object

Start by creating a new src function to represent the backend. Assuming we're going to create a src for mssql, you'd call it `src_mssql()`, and you'd follow the pattern of an existing src. A simplified version of `src_postgres()` is show below:

```{r, eval = FALSE}
src_postgres <- function(dbname = NULL, host = NULL, port = NULL, user = NULL,
                         password = NULL, ...) {

  con <- DBI::dbConnect(PostgreSQL(), host = host %||% "", dbname = dbname %||% "",
    user = user, password = password %||% "", port = port %||% "", ...)

  src_sql("postgres", con)
}
```

Use `src_sql()` to create a new S3 object with the correct structure. It must have a DBI connection, but it can store anything else that might be useful.

Next, implement a method for `brief_desc()` that briefly describe the source:

```{r}
#' @export
brief_desc.src_postgres <- function(x) {
  info <- DBI::dbGetInfo(con)
  host <- if (info$host == "") "localhost" else info$host

  paste0("postgres ", info$serverVersion, " [", info$user, "@",
    host, ":", info$port, "/", info$dbname, "]")
}
```

If you read the source code for the real `src_postgres()` you'll notice that it caches the `getGetInfo()` field on creation, since this saves a little time when printing tbls.

Before continuing, check that you can create a connection to a local database. Don't worry if the object doesn't print correctly, you'll need to add a few other methods first.

## Database methods

Next implement the database level methods in `dbi-s3.r` that need special behaviour for your database:

* `dbi_connect()`
* `db_list_tables()`
* `db_has_table()`
* `db_data_type()`

These should be very straightforward if there's an existing DBI interface, and only a little harder if not. This methods basically provide a shim around individual database packages to make them behave the same.

Once you have implemented these methods you should be able to connect to a local database get a useful description, including a list of tables. Don't continue on until you have that working.

## tbl

Next implement the `tbl()` method for your data source. This will probably just be

```{r, eval = FALSE}
tbl.src_mssql <- function(src, from, ...) {
  tbl_sql("mssql", src = src, from = from, ...)
}
```

Next implement the query related methods in `dbi-s3.r`:

* `qry_fields()`
* `table_fields()`
* `qry_fetch()`
* `qry_explain()`

These aren't documented, but are they are each fairly simple and there are lots of existing examples.

Before continuing, make sure you can connect to an existing table, and get a reasonable print out.

## ANSI SQL compliance

If you the database is not ANSI SQL compliant, you'll need to provide
methods for the sql level methods in `dbi-s3.r`:

* `sql_begin_trans()`
* `sql_commit()`
* `sql_insert_into()`
* `sql_create_indexes()`
* `sql_analyze()`

At this point, all the basic verbs (`summarise()`, `filter()`, `arrange()`, `mutate()` etc) should work.

## Lahman

To test your implementation, add `lahman_src()` and a line to `has_lahman()`. This will check that you've correctly implemented the sql modification functions, and is useful for testing other code.

## sql translation

To finish off, you can add custom R -> SQL translation by providing
a method for translate_env. This function should return an object
created by sql_variant(). See existing methods for examples.

If you're familiar with how your database works, and at least one
other database that dplyr already supports, this should be reasonably
simple, and I'm happy to help if you get stuck.  It's also possible
that a new database backend may need new methods - I'm also happy to
add those as needed.
