---
title: "Non-standard evaluation"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Non-standard evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
```

This vignette describe how non-standard evaluation (NSE) works in dplyr, and how the standard evaluation alternatives work.  The majority of functions in dplyr use NSE, but every function that uses NSE provides a version that only does standard evaluation (SE) ending with `_`.

```{r, message = FALSE}
library(dplyr)
# Use tbl_dt until C++ backend updated
mtcars <- tbl_dt(mtcars)

summarise(mtcars, mean(mpg))
summarise_(mtcars, ~mean(mpg))
summarise_(mtcars, .dots = list(~mean(mpg)))
```

Behind the scenes, NSE is powered by the [lazyeval](https://github.com/hadley/lazyeval) package. That means in the SE versions of accept a variety of inputs:

```{r}
library(lazyeval)

# Formulas
summarise_(mtcars, ~mean(mpg))
# Lazy objects
summarise_(mtcars, lazy(mean(mpg)))
# Quoted calls
summarise_(mtcars, quote(mean(mpg)))
# Strings
summarise_(mtcars, "mean(mpg)")
```

It's important to understand that every argument has two important components:

* A expression, e.g. `mean(mpg)`
* The environment in which to evaluate the expression.

Lazy captures both of these components. Formulas are particularly useful for SE because they capture both the expression and the environment, and the function name is very short!

What if you need to mingle constants and variables? Use the handy `lazyeval::interp()`:

```{r}
# Interp works with formulas, lazy objects, quoted calls and strings
interp(~ x + y, x = 10)
interp(lazy(x + y), x = 10)
interp(quote(x + y), x = 10)
interp("x + y", x = 10)

# Use as.name if you have a character string that gives a
# variable name
interp(~ mean(var), var = as.name("mpg"))
# or supply the quoted name directly
interp(~ mean(var), var = quote(mpg))

# Or a function!
interp(~ f(a, b), f = as.name("+"))
# Remember every action in R is a function call:
# http://adv-r.had.co.nz/Functions.html#all-calls

# If you've built up a list of values through some other
# mechanism, use .values
interp(~ x + y, .values = list(x = 10))

# You can also interpolate variables defined in the current
# environment, but this is a little risky.
y <- 10
interp(~ x + y, .values = environment())
```
