---
title: "Programming with dplyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Programming with dplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
set.seed(1014)
```

## Introduction

Most dplyr verbs use **tidy evaluation** in some way. Tidy evaluation is a special type of non-standard evaluation used throughout the tidyverse. There are two basic forms found in dplyr:

* `arrange()`, `count()`, `filter()`, `group_by()`, `mutate()`, and 
  `summarise()` use **data masking** so that you can use data variables as if 
  they were variables in the environment (i.e. you write `my_variable` not
  `df$myvariable`). 

* `across()`, `relocate()`, `rename()`, `select()`, and `pull()` use 
  **tidy selection** so you can easily choose variabe based on their position, 
  name, or type (e.g. `starts_with("x")` or `is.numeric`).

Data masking and tidy selection make interactive data exploration fast and fluid, but add some new challenges when you attempt to use them indirectly, as in a for loop or function. This vignette shows you how to overcome those challenges. We'll first go over the basics of data masking and tidy selection, talk about how to use them indirectly, and then show you a number of recipes to solve common problems.

This vignette will give you the minimum knowledge you need to be an effective programmer with tidy evaluation. If you'd like to learn more about the underlying theory, or how precisely its different from non-standard evaluation, we recommend that you read the Metaprogramming chapters in [_Advanced R_](https://adv-r.hadley.nz).

```{r, message = FALSE}
library(dplyr)
```

## Data masking

Data masking makes data manipulation faster because it requires less typing. In most (but not all[^subset]) base R functions you need to refer to variables with `$`, leading to code that repeats the name of the data frame many times:

```{r, results = FALSE}
starwars[starwars$homeworld == "Naboo" & starwars$species == "Human", ,]
```

[^subset]: dplyr's `filter()` is inspired by base R's `subset()`. `subset()` provides data masking, but not with tidy evaluation, so the techniques described in this chapter don't apply to it.

The dplyr equivalent of this code is more concise because you only need to write `starwars` once:

```{r, results = FALSE}
starwars %>% filter(homeworld == "Naboo", species == "Human")
```

### Data- and env-variables 

The key idea behind data masking is that it blurs the line between the two different meanings of the word "variable":

* **env-variables** are "programming" variables that live in an environment.
  They are usually created with `<-`. 

* **data-variables** are "statistical" variables that live in a data frame.
  They usually come from data files (e.g. `.csv`, `.xls`), or are created 
  manipulating existing variables. 
  
To make those definitions a little more concrete, take this piece of code:

```{r}
df <- data.frame(x = runif(3), y = runif(3))
df$x
```

It creates a env-variable, `df`, that contains two data-variables, `x` and `y`. Then it extracts the data-variable `x` out of the env-variable `df` using `$`.

I think this blurring of the meaning of variable is a really nice feature for interactive data analysis, because it allows you to refer to data-vars as is, without any prefix. And this seems to be fairly intuitive, since many newer R users will attempt to write `diamonds[x == 0 | y == 0, ]`. 

Unfortunately this benefit does not come for free. When you start to program with these tools, you're going to have to grapple with the distinction. And this will be hard because you've never had to think about it before, so it'll take a while for your brain to learn these new concepts and categories. However, once you've teased apart the idea of "variable" in data-variable and env-variable, I think you'll find it fairly straightforward to use.

### Indirection

The main challenge of programming with function that use data masking arises when you introduce some indirection, i.e. instead of you directly typing the name of a data-variable you want to get it from an env-variable. There are two main cases:

*   When you have the data-variable in a function argument (i.e. an env-variable
    that holds a promise[^promise]), you need to **embrace** the argument by
    surrounding it in doubled parentheses, like `filter(df, {{ var }})`.
    
    The following function uses embracing to create a wrapper around 
    `summarise()` that computes the minimum and maximum values of a variable,
    as well as the number of observations that were summarised.

    ```{r, results = FALSE}
    var_summary <- function(df, var) {
      df %>%
        summarise(n = n(), min = min({{ var }}), max = max({{ var }}))
    }
    mtcars %>% 
      group_by(cyl) %>% 
      var_summary(mpg)
    ```

*   When you have an env-variable that is a character vector, you need to index 
    into the `.data` pronoun with `[[`, like 
    `summarise(df, mean = mean(.data[[var]]))`.
    
    The following example uses `.data` to count the number of unique values in
    each variable of `mtcars`:

    ```{r, results = FALSE}
    for (var in names(mtcars)) {
      mtcars %>% count(.data[[var]]) %>% print()
    }
    ```


[^promise]: In R, arguments are lazily evaluated which means that until you attempt to use, they don't hold a value, just a __promise__ that describes how to compute the value. You can learn more at <https://adv-r.hadley.nz/functions.html#lazy-evaluation>

## Tidy selection

Data masking makes it easy to compute on values within a dataset. Tidy selection is a complementary tool that makes it easy to work with the columns of a dataset.

### The tidyselect DSL

Underneath all functions that use tidy selection is the [tidyselect](http://tidyselect.r-lib.org/) package. It provides a miniature domain specific language that makes it easy to select columns by name, position, or type. For example:

* `select(df, 1)` selects the first column; `select(1, last_col())` selects
  the last column.
  
* `select(df, c(a, b, c))` selects columns `a`, `b`, and `c`.

* `select(df, starts_with("a"))` selects all columns whose name starts with "a";
  `select(df, ends_with("a"))` selects all columns whose name ends with "z".
  
* `select(df, is.numeric)` selects all numeric columns.

(See more details in `?dplyr_tidy_select`).

### Indirection selection

As with, data masking there are two forms of indirection. 

*   When you have the data-variable in an env-variable that is a functon
    argument, you use the same technique as data masking: you **embrace** the
    argument by surrounding it in doubled parentheses.
    
    The following function summarises a data frame by computing
    the mean of all variables selected by the user.

    ```{r, results = FALSE}
    summarise_mean <- function(df, vars) {
      df %>% summarise(n = n(), across({{ vars }}, mean))
    }
    mtcars %>% 
      group_by(cyl) %>% 
      summarise_mean(is.numeric)
    ```

*   When you have an env-variable that is a character vector, you need to use
    `all_of()` or `any_of()` depending on whether or not you want the
    function to error if a variable is not found. 
    
    The following code uses `all_of()` to select all of the variables found
    in a character vector; then `!` plus `all_of()` to select all of the 
    variables *not* found in a character vector.

    ```{r, results = FALSE}
    vars <- c("mpg", "vs")
    mtcars %>% select(all_of(vars))
    mtcars %>% select(!all_of(vars))
    ```

## How tos

The examples here are somewhat inauthentic because we've reduced them down to very simple components to make them easier to understand. They're so simple that you might wonder why we bother writing a function at all. But it's a good idea to learn the ideas on simple examples, so that you're better prepared to apply them to the more complex situations you'll see in your own code.

### How to tell if a function uses tidy evaluation

Tidy evaluation is a property that applies not to an entire function, but to specific arguments of a function. You can tell if an argument uses special evaluation because its documentation will have special a label that looks like `<tidy-eval>`.

### Function with user-supplied data

If you read the documentation for the dplyr verbs, you'll see that they don't use tidy evaluation for the `.data` argument. This makes it simple to reduce duplication when repeated code only varies the data:

```{r, eval = FALSE}
mutate(df1, y = a + x)
mutate(df2, y = a + x)
mutate(df3, y = a + x)
mutate(df4, y = a + x)
```

```{r}
mutate_y <- function(df) {
  mutate(df, y = a + x)
}
```

### Function with a user-supplied expression

What about when a repeated code varies over an argument that uses tidy evaluation?

```{r}
starwars %>%
  group_by(species) %>%
  summarise(mean = mean(mass, na.rm = TRUE))

starwars %>%
  group_by(homeworld) %>%
  summarise(mean = mean(mass, na.rm = TRUE))
```

You'll need to embrace that argument with `{{ }}`

```{r}
my_summarise <- function(df, group_var) {
  df %>%
    group_by({{ group_var }}) %>%
    summarise(mean = mean(mass))
}

my_summarise(starwars, species)
```

This generalises in a straightforward way if you want to use one user supplied expression in multiple places. For example, if you wanted to write a function to reduce the duplication here:

```{r}
df <- tibble(a = 1:10, b = runif(10))

df %>% summarise(mean = mean(a), sum = sum(a), n = n())
df %>% summarise(mean = mean(a * b), sum = sum(a * b), n = n())
```

You'd just use:

```{r}
my_summarise2 <- function(df, expr) {
  expr <- enquo(expr)

  df %>% summarise(
    mean = mean({{ expr }}),
    sum = sum({{ expr }}),
    n = n()
  )
}
df %>% my_summarise2(a)
df %>% my_summarise2(a * b)
```

### Function with multiple user-supplied expressions

If you just want to pass expressions along without transforming them, you can use `...`. This is most often useful when you want to give the user full control over a single part of the pipeline, like a `group_by()` or a `mutate()`.

```{r}

my_summarise <- function(df, ...) {
  df %>%
    group_by(...) %>%
    summarise(mass = mean(mass, na.rm = TRUE), height = mean(height, na.rm = TRUE))
}

my_summarise(starwars, homeworld)
```
  
### Function with multiple user-supplied variables

If you want the user to provide a set of variables that are then transformed, use `across()`:

```{r}
my_summarise <- function(df, summary_vars) {
  df %>%
    summarise(across({{ summary_vars }}, ~ mean(., na.rm = TRUE)))
}
starwars %>% 
  group_by(species) %>% 
  my_summarise(c(mass, height))
```

You can use this same idea for multiple sets of input variables:

```{r}
my_summarise <- function(df, group_by, summarise_by) {
  df %>%
    group_by(across({{ group_by }})) %>% 
    summarise(across({{ summarise_by }}, mean))
}
```

### Loop over multiple variables

If you want to loop over variables you need a slightly different approach, because precisely storing expressions in a vector requires greater knowledge of the underlying theory. A simpler approach is to put the variable names in a character vector and use the special `.data` pronoun:

```{r, eval = FALSE}
for (var in names(mtcars)) {
  mtcars %>% count(.data[[var]]) %>% print()
}
```

(Note that `.data` is not a data frame; it's a special construct that allows you to access the current variables either directly, with `.data$x` or indirectly with `.data[[var]]`. Don't expect other functions to work with it.)

### Use a variable from an Shiny input

Many Shiny input controls return character vectors, so you can use the same approach as above: `.data[[input$var]]`.

```{r, eval = FALSE}
library(shiny)
ui <- fluidPage(
  selectInput("var", "Variable", choices = names(diamonds)),
  tableOutput("output")
)
server <- function(input, output, session) {
  data <- reactive(filter(diamonds, .data[[input$var]] > 0))
  output$output <- renderTable(head(data()))
}
```

See <https://mastering-shiny.org/action-tidy.html> for more details and case studies.

### Eliminating the `R CMD check` `NOTE`

```{r}
my_summary_function <- function(data) {
  data %>% 
    filter(x > 0) %>% 
    group_by(grp) %>% 
    summarise(y = mean(y), n = n())
}
```

```
N  checking R code for possible problems
   my_summary_function: no visible binding for global variable ‘x’, ‘grp’, ‘y’
   Undefined global functions or variables:
     x grp y
```

```{r}
#' @importFrom rlang .data
my_summary_function <- function(data) {
  data %>% 
    filter(.data$x > 0) %>% 
    group_by(.data$grp) %>% 
    summarise(y = mean(.data$y), n = n())
}
```
