---
title: "Programming with dplyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Programming with dplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
library("dplyr")
library("purrr")
library("rlang")
set.seed(1014)
```


Programming with dplyr requires a bit of special knowledge because most dplyr
verbs are not normal functions. They are **quoting functions**. In this vignette
you will learn about quoting functions, what challenges they pose for
programming, and how **tidy evaluation** solves those problems.


## Introduction

### Regular functions versus quoting functions

R functions can be categorised in two broad categories: regular functions and
quoting functions. These functions differ in the way they get their arguments.
Regular functions only see values. It does not matter what the expression
supplied as argument is or which variables it involves. The value is computed
following the standard rules of evaluation [^1]. The fundamental regular
function is `identity()`, it returns the value of its argument. Because only the
value matters, all of these statements are completely equivalent:

```{r}
identity(6)

identity(2 * 3)

a <- 2
b <- 3
identity(a * b)
```

On the other hand, a quoting function sees the *expression* typed as argument
rather than the value of this expression. The expression might be evaluated a
bit later or might not be evaluated at all. The fundamental quoting function is
`quote()`, it returns the expression ot its argument:

```{r}
quote(6)

quote(2 * 3)

quote(a * b)
```

In fact the action of quoting is something that all programmers are familiar
with because this is they create strings. `"` is a quoting operator. It is a
signal that the supplied characters are not code but text. As an R programmer,
you are also probably familiar with the formula operator `~`. This quoting
operator returns one or two quoted expressions. Thus the three following
expressions are doing something similar, they are quoting their input:

```{r}
"a * b"

~a * b

quote(a * b)
```

The first statement returns a quoted string and the other two return quoted code
in a formula or as a bare expression.

[^1]: This is why regular functions are said to use standard evaluation unlike
      quoting functions which use non-standard evaluation (NSE).


### Changing the context of evaluation

A quoted expression can be **evaluated** using the function `eval()`. Let's
quote an expression that represents the subset of lowercase letters from 1 to 5,
and evaluate this:

```{r}
x <- quote(letters[1:5])

x

eval(x)
```

Of course this is not very impressive, you could just type the expression
normally to get this value. But one of R's most important feature is that you
can change the context of evaluation to obtain different results. A context,
also called **environment**, is basically a set that links symbols to
values. The namespaces of packages are such context. For instance, in the
context of the base namespace, the symbol `letters` is given the value of a
character vector of lowercase letters. However it could mean something different
in another context. We could create a context where `letters` represent the
uppercase letters in reverse order! Evaluating a quoted expression in such a
context could return a completely different result:

```{r}
context <- list(letters = rev(LETTERS))

x

eval(x, context)
```

Interestingly, data frames can be used as evaluation contexts. In a data frame
context, the column names represent vectors so that you can refer to those
columns in an expression:

```{r}
data1 <- tibble(mass = c(70, 80, 90), height = 1.6, 1.7, 1.8)
data2 <- tibble(mass = c(75, 85, 95), height = 1.5, 1.7, 1.9)

bmi_expr <- quote(mass / height^2)

eval(bmi_expr, data1)

eval(bmi_expr, data2)
```

In the last snippet we are creating an expression with `quote()` and we evaluate
it manually with `eval()`. However quoting functions typically perform the
quoting and the evaluation for you behind the scene:

```{r}
with(data1, mass / height^2)

with(data2, mass / height^2)
```

For this reason quoting functions usually take a data frame as input in addition
to user expressions so they can be evaluated in the context of the data. This is
a powerful feature that gives R its identity as a data-oriented programming
language. Quoting functions are everywhere in R:

*   `with(data, expr)` evaluates `expr` in the context of `data`.

*   `lm(formula, data)` creates a design matrix with predictors evaluated in the
    context of `data`.

*   `mutate(data, new = expr)` creates a `new` column from an expression evaluated
    in the context of `data`.

*   `ggplot(data, aes(expr))` defines the `x` aesthetic as the value of `expr`
    evaluated in the context of `data`.

In the context of the dplyr interface, quoting the arguments has two benefits:

*   Operations on data frames can be expressed succinctly because
    you don't need to repeat the name of the data frame. For example,
    you can write `filter(df, x == 1, y == 2, z == 3)` instead of
    `df[df$x == 1 & df$y ==2 & df$z == 3, ]`.

*   dplyr can choose to compute results in a different way to base R.
    This is important for database backends because dplyr itself doesn't
    do any work, but instead generates the SQL that tells the database
    what to do.

Unfortunately the benefits of quoting functions do not come for free. While they
simplify direct inputs, they make it harder to program the inputs. Quoting works
for you when *you* use dplyr but works against you when *your functions* use
dplyr.


### Varying quoted inputs

The issue of referential transparency to do with the difficulty of passing
contextual variables in order to vary the inputs of quoting functions. When you
pass variables to quoting functions they get quoted along with the rest of the
expression.

To see the problem more clearly, let's define a simple quoting function [^2]
that pastes its inputs as a string:

```{r}
cement <- function(..., .sep = " ") {
  strings <- map(exprs(...), as_string)
  paste(strings, collapse = .sep)
}
```

Compared to the regular function `paste()`, the quoting function `cement()`
saves a bit of typing because it performs the string-quoting automatically:

```{r}
paste("it", "is", "rainy")

cement(it, is, rainy)
```

Now what if we wanted to store the weather adjective in a variable? `paste()`
has no issue on that front because it gets the value of the argument rather than
its expression. On the other hand if we pass a variable to `cement()`, it would
be quoted just like the other inputs and `cement()` would never get to see its
contents:

```{r}
x <- "shiny"

paste("it", "is", x)

cement(it, is, x)
```

The solution to this problem is a special syntax that signals the quoting
function that part of the argument is to be unquoted, i.e., evaluated right
away. The ability to mix quoting and evaluation is called **quasiquotation** and
is the main tidy eval feature.

[^2]: As we will see later on `exprs()` captures the expressions of its inputs.
      Passing `...` to `exprs()` returns a list of quoted arguments forwarded
      through dots.


## Quasiquotation

> Put simply, quasi-quotation enables one to introduce symbols that stand for
> a linguistic expression in a given instance and are used as that linguistic
> expression in a different instance.
--- [Willard van Orman Quine](https://en.wikipedia.org/wiki/Quasi-quotation)

As we have seen, automatic quoting makes R and dplyr very convenient for
interactive use but makes it difficult to refer to variable inputs. The solution
to this problem is __quasiquotation__, which allows you to evaluate directly
inside an expression that is otherwise quoted. Quasiquotation was coined by
Willard van Orman Quine in the 1940s, and was adopted for programming by the
LISP community in the 1970s. Quasiquotation is available (or will soon be) in
all quoting functions of the tidyverse thanks to the tidy evaluation framework.


### The bang! bang! operator

The tidy eval syntax for unquoting is `!!`. Anything supplied to to this
operator is evaluated right away and the result is substituted in place. Let's
see `!!` in action in our `cement()` function:

```{r}
x <- "shiny"

cement(it, is, !! x)
```

Even though the arguments are quoted, `!! x` signals that `x` should be
evaluated right away. From `cement()` perspective, it's as if the user had typed
`"shiny"` instead of `!! x`.

We have seen above that the fundamental quoting function in base R is `quote()`.
In the tidyverse, it is `expr()`. All it does is to quote its argument with
quasiquotation support and returns it right away:

```{r}
expr(x)

expr(!! x)
```

`expr()` is especially useful for debugging quasiquotation. You can wrap it
around any expression in which you use `!!` to examine the effect of unquoting.
Let's try it with `cement()`:

```{r}
expr(cement(it, is, !! x))
```

This technique is essential to work your way around to mastering tidy eval.


### Creating symbols

Now that we are armed with quasiquotation, let's try to program with the dplyr
verb `mutate()`. We'll take a BMI computation as running example.

```{r}
# Rescale height
starwars <- mutate(starwars, height = height / 100)

transmute(starwars, bmi = mass / height^2)
```

Let's say we want to vary the height input. A first intuition might be to store
the column name in a variable and unquote it. But we get an error:

```{r, error = TRUE}
x <- "height"

transmute(starwars, bmi = mass / (!! x)^2)
```

The error message indicates a type error. A binary operator expected a numeric
input but got something else. The error becomes clear if we use `expr()` to
debug the unquoting:

```{r}
expr(transmute(starwars, bmi = mass / (!! x)^2))
```

We are unquoting a string and that's exactly what `transmute()` uses to evaluate
the BMI. This can't work! We need to unquote something that looks like code
instead of a string. What we are looking for is a **symbol**. A symbol is a
string that references an object in a context. Symbols are the meat of R
code. In `foo(bar)`, `foo` is a symbol that references a function and `bar` is a
symbol that references some object.

There are two ways of creating symbolic R code objects: by quotation or by
construction. We already know how to create symbols by quoting. However that
does not help us much because we face the same issue again, namely that the
quoted symbol is a constant that can't be varied:

```{r}
quote(height)

expr(height)
```

The other way is to build it out of a string using the constructor `sym()`.
Constructors are regular functions and can be programmed with variables:

```{r}
sym("height")

x <- "height"
sym(x)
```

Let's build a symbol and try to unquote it in the transmute expression. Using
`expr()` to examine the effect of unquoting, things are looking good:

```{r}
x <- sym("height")

expr(transmute(starwars, bmi = mass / (!! x)^2))
```

And indeed it now works!

```{r}
transmute(starwars, bmi = mass / (!! x)^2)
```


## Creating a wrapper around a dplyr pipeline

Quasiquotation is all we need to write our first wrapper function around a dplyr
pipeline. The goal is to write reliable functions that reduce duplication in
our data analysis code. Let's say that we often take a grouped average using
dplyr and our scripts are littered with little pipelines that look like this:

```{r}
starwars %>%
  group_by(species) %>%
  summarise(avg = mean(height))
```

It would be a good idea to extract this logic into a function. It would reduce
the risk of writing a typo and would make our code more concise as well as
clearer if we choose a good name for this function.

We know from the previous sections that this kind of naive wrapper will not
work because the variable names will be automatically quoted:

```{r, error = TRUE}
mean_by <- function(data, var, group) {
  data %>%
    group_by(group) %>%
    summarise(avg = mean(var))
}

mean_by(starwars, "species", "height")
```

* In the best case the column names they contain will be ignored. For instance `group_by()`
  looks for a column named `group` and doesn't see the string `"species"`.

* In the worst case they will be misused. For instance `summarise()` would try
  to take the average of the string `"height"`.

To avoid this, our wrapper simply needs to construct symbols from its inputs and
unquote them in the pipeline:

```{r}
mean_by <- function(data, var, group) {
  var <- sym(var)
  group <- sym(group)

  data %>%
    group_by(!! group) %>%
    summarise(avg = mean(!! var))
}

mean_by(starwars, "height", "species")

mean_by(starwars, "mass", "eye_color")
```


### Creating your own quoting functions

The wrapper that we just created is a regular function that takes strings and
doesn't quote any of its inputs. This has the advantage that it is easy to
program with but the inconvenient that it doesn't integrate well with the rest
of the tidyverse verbs. Fortunately it is easy to transform the wrapper into a
quoting function.

First we need to choose which of our wrapper arguments should be quoted. Given
the friction that quotation causes for programming, it is best to only quote
arguments when absolutely necessary, i.e. when it makes sense to refer to data
frame columns. In dplyr, the argument that takes a data frame (which is always
the first argument in order to be compatible with pipes) is never quoted. We'll
apply the same logic to our wrapper and only quote the `group` and `var`
arguments.

Tidy eval provides two functions to quote an argument supplied by the caller of
a function. Both of those enable quasiquotation:

* `enexpr()` which returns a raw expression.
* `enquo()` which returns an expression wrapped in a **quosure**.

Let's first try `enexpr()` in a simple function that does nothing but capture
its argument and return it right away:

```{r}
quoting <- function(x) enexpr(x)

x <- sym("foo")

quoting(x)

quoting(!! x)
```

We have in fact just reinvented `expr()`! Indeed `expr()` is a simple wrapper
around `enexpr()`:

```{r}
dplyr::expr
```

In the same vein, `quo()` is a wrapper around `enquo()`. All it does is to
capture the expression of its argument, store it in a quosure, and return it as
is:

```{r}
dplyr::quo

quo(x)

quo(!! x)
```

A quosure is like a raw expression except that it is evaluated in the original
context of its capture. It combines an expression (a quote) and a context (an
enclosure) in a single object. We'll see below why it is important to keep track
of the original context of arguments. For now, let's just use it in our pipeline
wrapper to transform it into a quoting function. As a reminder here is the
current definition of our function:

```{r}
mean_by <- function(data, var, group) {
  var <- sym(var)
  group <- sym(group)

  data %>%
    group_by(!! group) %>%
    summarise(avg = mean(!! var))
}
```

All we need to do is to replace the `sym()` constructor by `enquo()`:

```{r}
mean_by <- function(data, var, group) {
  var <- enquo(var)
  group <- enquo(group)

  data %>%
    group_by(!! group) %>%
    summarise(avg = mean(!! var))
}
```

The wrapper now automatically quotes its arguments. This has several
implications:

*   First the user no longer has to supply quoted strings:

    ```{r}
    mean_by(starwars, height, species)
    ```

*   Secondly, while `sym()` assumed that the supplied arguments were symbols,
    `enquo()` captures arbitrary expressions. This is a good fit for our wrapper
    because both `group_by()` and `summarise()` accept complex expressions:

    ```{r}
    mean_by(starwars, height * 100, as.factor(species))
    ```

*   Since our function now quotes its arguments, it is no longer programmable in
    the usual way. If another function passes variables to `mean_by()`, it needs
    to use quasiquotation itself. A typical composition of quoting functions
    thus looks like a chain of quoted and unquoted arguments:

    ```{r}
    mean_by_species <- function(data, var) {
      var <- enquo(var)
      mean_by(data, !! var, species)
    }

    mean_by_species(starwars, height)
    ```

Thanks to `enquo()` we now have a wrapper function that quotes its inputs and
interacts with dplyr verbs via quasiquotation. It is getting pretty close to a
real tidyverse-like user interface! However we could still improve a few things,
like the automatic labelling of column names which could be better. It would
also be nice if the wrapper could accept a variable number of arguments like
other dplyr or tidyr verbs. We'll address the latter issue first.


### Accepting multiple arguments

Whether our wrapper should take multiple grouping variables or multiple
variables to average is a design decision that could go either way depending on
your needs. In this tutorial we'll allow multiple grouping variables.

It is relatively easy to write R functions that accept an unspecified number of
arguments. The function just takes `...` as argument. In the body of the
function `...` are then forwarded to another variadic function that is in charge
of materialising the arguments. The end point is typically the list function:

```{r}
variadic <- function(...) list(...)

variadic("foo", "bar")
```

Passing on arguments through dots to quoting functions is very easy. Unlike
named arguments which need to be repeatedly quoted and unquoted, the `...`
object can just be passed along:

```{r}
mean_by <- function(data, var, ...) {
  var <- enquo(var)

  data %>%
    group_by(...) %>%
    summarise(avg = mean(!! var))
}
```

Your users can now create grouped averages for any combination of groups!

```{r}
mean_by(starwars, height, species, eye_color)
```

You can learn about more advanced ways of dealing with multiple arguments with
`exprs()`, `quos()` and `syms()` in the section on variadic quasiquotation
below.


### Labelling inputs

dplyr functions try their best to provide useful column names for new columns.
This is an area where our wrapper could use some improvement:

```{r}
names(mean_by(starwars, height, as.factor(species)))
```

First note that the issue is in fact already solved for the grouping
variables. That's a benefit from taking arguments with `...`, they accept
optional names:

```{r}
mean_by(starwars, height, species_fct = as.factor(species))
```

However for named arguments we need to do a bit more work. We'll make use of two
tidy eval features:

*   `quo_name()` which is a helper that transforms an arbitrary expression
    (including quosures) to a name that is suitable for data frames:

    ```{r}
    wrapper <- function(x) {
      x <- enquo(x)
      quo_name(x)
    }

    wrapper(foo)

    wrapper(foo(bar, baz()))
    ```

*   The `:=` operator. It makes it possible to unquote on the left-hand side of
    an argument. Since the LHS of `=` is automatically quoted, it makes sense to
    have quasiquotation for argument names:

    ```{r}
    x <- "Column Name"
    summarise(starwars, !! x := n())
    ```

We can give a nice default name to the column of averages by transforming the
captured variable to a name and pasting a prefix at its front:

```{r}
mean_by <- function(data, var, ...) {
  var <- enquo(var)

  name <- quo_name(var)
  name <- paste0("avg_", name)

  data %>%
    group_by(...) %>%
    summarise(!! name := mean(!! var))
}
```

We get a good name that reflects the user input, even when the argument is a
complex expression:

```{r}
mean_by(starwars, height, species)

mean_by(starwars, identity(height), species)
```

Overall the most flexible interface is `...` since they let the user specify
custom names. But what if we want to add a prefix to the grouping variables as
well? Then we can't just pass the `...` variable down to `group_by()`, we have
to capture all the variables in the dots and modify their names before passing
them on. This calls for more advanced means of working with multiple arguments.


### Capturing and modifying arguments in `...`

Up until now, we have captured *named arguments* with `enquo()`, we have
forwarded variadic arguments by passing `...` to tidy eval functions, but we
have yet to actually capture those arguments contained in `...`. Getting a hold
on the expressions supplied as `...` arguments is necessary in order to make
modifications such as changing the argument names.

As we have seen arguments transiting through dots need to be materialised with
endpoint functions such as `c()` or `list()`. Tidy eval provides two variadic
endpoints for dots: `exprs()` and `quos()`. These functions quote all of their
inputs and return them in a list of expressions or quosures:

```{r}
exprs(foo, bar)

quos(baz, bam)
```

Thanks to the magic of `...` forwarding, `exprs()` and `quos()` will capture all
arguments passed through dots:

```{r}
quoting <- function(...) {
  exprs(foo, ...)
}

quoting(bar(baz))
```

We'll first experiment with a simple `group_by()` wrapper before applying our
new knowledge to the `mean_by()` wrapper. This wrapper will prefix all grouping
variables with `grp_`. To achieve this there are two problems to solve:
modifying the names, and forwarding the list of captured arguments to
`group_by()` once we are done changing the names. Let's start this function with
a bare skeleton. It will take a data frame, a prefix for the group names, and an
undefined number of grouping arguments:

```{r}
prefixed_group_by <- function(data, prefix, ...) {
  groups <- quos(...)
  groups
}

groups <- prefixed_group_by(starwars, "grp_", as.factor(species), color = eye_color)

groups

names(groups)
```

We have supplied two arguments as grouping variable. The first is an unnamed
complex expression, the second is a named symbol. The first thing to do is to
give a default name to arguments. One way to obtain default names would be to
map `quo_name()` over the relevant elements but there is an easier way. `quos()`
will do it for you if you switch on the `.named` argument:

```{r}
quos <- quos(foo(bar), baz = foo(), .named = TRUE)
names(quos)
```

We are now in a good position for adding a prefix to the names of captured
arguments:

```{r}
prefixed_group_by <- function(data, prefix, ...) {
  groups <- quos(..., .named = TRUE)
  names(groups) <- paste0(prefix, names(groups))
  groups
}

groups <- prefixed_group_by(starwars, "grp_", as.factor(species), color = eye_color)
names(groups)
```

Alright! We only have one last problem to solve. We need a way to forward this
list of arguments to `group_by()`. Unquoting the list with `!!` is not helpful
here because `group_by()` expects separate arguments and wouldn't know what to
do with a whole list. This leads us to `!!!`, one of the most handy features of
tidy eval.


### Unquote-splicing arguments with `!!!`

The __unquote-splicing__ operator `!!!` is a variant of simple unquoting. Just
like `!!`, it evaluates its right-hand side right away. The difference is in the
way it substitutes the result in the surrounding call:

*   `!!` substitutes in place:

    ```{r}
    expr(call(!! 1:5))
    ```

*   `!!!` takes a vector and substitutes all its elements in the call:

    ```{r}
    expr(call(!!! 1:5))
    ```

This is exactly what we need to forward a list of captured arguments to
`group_by()`!

```{r}
prefixed_group_by <- function(data, prefix, ...) {
  groups <- quos(..., .named = TRUE)
  names(groups) <- paste0(prefix, names(groups))

  group_by(data, !!! groups)
}

prefixed_group_by(starwars, "grp_", as.factor(species), color = eye_color)
```

Modifying `mean_by()` to automatically prefix the grouping factors is now
child's play:

```{r}
mean_by <- function(data, var, ...) {
  var <- enquo(var)

  name <- quo_name(var)
  name <- paste0("avg_", name)

  data %>%
    prefixed_group_by("grp_", ...) %>%
    summarise(!! name := mean(!! var))
}

mean_by(starwars, height, species, eye = eye_color)
```


### Wrapping it up

In order to write our little wrapper, we have learned to:

*   Quote R code with `quote()` and `expr()` and construct symbols with `sym()`.

*   Capture named arguments with `enquo()` and `...` arguments with `quos()`.

*   Unquote single arguments with `!!` and multiple arguments with `!!!`.

*   Use `:=` to enable `!!` on the left-hand side of a named argument.

*   Debug the unquoting by wrapping `expr()` around an expression.

*   Use `quo_name()` and `quos(.named = TRUE)` to provide default names to
    captured arguments.

This set of techniques will get you a long way as quasiquotation is really the
meat of programming with tidy eval. `enquo()` and `quos()` return quosures that
are more reliable than bare expressions but you don't have to understand how
quosures work or why they are needed to effectively use tidy eval.

When you feel ready, you can learn about the concept of quosures. It will
improve your understanding of R programming and you will gain knowledge that can
be applied to R functions. Quosures and closures (the technical name of R
functions) have a lot in common!


## Where do quoting verbs find things?

### Contexts and hierarchical ambiguity

### Solving ambiguity with quasiquotation and the `.data` pronoun

### Raw expressions versus contextual expressions (quosures)
