---
title: "Programming with dplyr"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Non-standard evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(dplyr)
```

The key dplyr verbs take their arguments by expression rather
than by value. This means that instead of evaluating the arguments working with their values, they capture the expression, you typed. This has two main benefits:

*   Operations on data frames can be expressed more succinctly because
    in expressions like `filter(df, x == 1, y == 2, z == 3)` dplyr
    assumes that `x`, `y`, and `z` are variables in df.
    
*   dplyr doesn't need to compute the results as same way as base R.
    For example, for remote tables, dplyr doesn't do any computation
    but instead generates SQL that causes the remote database to do 
    the computation.

Unfortunately these benefits do not come for free. There are two main costs:

*   dplyr code is not referentially transparent. That means that you
    can not replace a variable with it's value. In other words,
    this code:
    
    ```{r, eval = FALSE}
    filter(df, my_var == 1)
    ```
    
    Is not equivalent to this code:
    
    ```{r, eval = FALSE}
    var <- my_var
    filter(df, var == 1)
    ```
    
    This makes it harder to create functions that replace 
    copied-and-pasted code.
    
*   dplyr code is ambiguous. Depending on what variables are
    defined where, `filter(df, x == y)` could be equivalent
    to any of:

    ```{r, eval = FALSE}
    df[df$x == df$y, ]
    df[df$x == y, ]
    df[x == df$y, ]
    df[x == y, ]
    ```
    
    This ambiguity makes functions that use dplyr more unpredictable
    than you might desire.
    
The goal of this vignette is to show you how to use dplyr's __quasiquotation__ and __pronouns__ to resolve both these issues, allowing to you write functions to reduce copy-and-paste. Rather than diving into the theory behind these ideas (called tidyeval), we'll start with a few recipes showing you how to reduce duplication in common copy and paste scenarios.

## Programming recipes

### Different data sets

```{r, eval = FALSE}
mutate(df1, y = a + x)
mutate(df2, y = a + x)
mutate(df3, y = a + x)
mutate(df4, y = a + x)

# OK, but if `a` or `x` is not in `df`, will look
# in environment
mutate_y <- function(df) {
  mutate(df, y = a + x)
}

# Use .data "pronoun" to ensure that mutate() only looks
# in data, not environment. IF `a` and `x` are not present 
# in `df`, then this will throw an error. Additionally, if
# this code is in a package, it avoids an R CMD CHECK note
# (you will need to import rlang::.data)
mutate_y <- function(df) {
  mutate(df, y = .data$a + .data$x)
}
```

### Different input variable

```{r, eval = FALSE}
summarise(df, mean(a))
summarise(df, mean(b))
summarise(df, mean(c))
summarise(df, mean(e ^ 2))

# Use !! to "unquote" expr, inserting it's value into
# the expression
summarise_mean <- function(df, expr) {
  summarise(df, mean(!!expr))
}
```

### Different input and output variable

```{r, eval = FALSE}
summarise(df, mean_a = mean(a))
summarise(df, mean_b = mean(b))
summarise(df, mean_c = mean(c))

# Use := to change the name of a function argument
# (necessary because !!x == y is not valid R code)
summarise_mean <- function(df, expr) {
  new_name <- paste0("mean_", expr_text(expr))
  summarise(df, !!new_name := mean(!!expr))
}

# Note that this will generate non-syntactic names
# if expr contains a space or other special character
```

### Different arguments

```{r, eval = FALSE}
summarise_mean(df, mean(a, na.rm = TRUE))
summarise_mean(df, mean(b))
summarise_mean(df, mean(c, trim = 0.2))

summarise_mean <- function(df, var, ...) {
  # Capture unevaluated dots
  dots <- rlang::quosures(...)
  summarise(df, mean(!!var, !!!dots))
}
```

## Quoting

Quoting is the action of capturing an expression instead of evaluating
it. All expression-based functions quote their arguments and get the R
code as an expression rather than the result of evaluating that
code. If you are an R user, you probably quote expressions on a
regular basis. One of the most important quoting operators in R is the
_formula_. It is famously used for the specification of statistical
models:

```{r}
disp ~ cyl + drat
```

The other quoting operator in base R is `quote()`. It returns a raw
expression rather than a formula:

```{r}
# Computing the value of the expression:
toupper(letters[1:5])

# Capturing the expression:
quote(toupper(letters[1:5]))
```

In practice, the formula is the most important way of quoting an
expression because it knows about the variables defined in the current
context. For example, if you create a formula in the global workspace,
it knows it was created there and that symbols may refer to variables
defined in that context rather than in the context of a function. For
instance, the variable `letters` refers to a standard R object that is
accessible from the global workspace. Typing `~letters` from that
workspace records both the symbol `letters` and the context where it
can be found.

In R and other programming languages, these contexts are called
_environments_. A formula bundles both an expression and an
environment. When an object keep tracks of an environment, it is said
to have an enclosure (functions are another example of such
objects). For this reason we use a special name to refer to one-sided
formulas. We call them _quosures_: one-sided formulas are quotes (they
carry an expression) that bundle an enclosure. Note that unlike
quosures, two-sided formulas bundle not one but two expressions, and
are thus slightly more complex objects. Here is our first quosure:

```{r}
~toupper(letters[1:5])
```

It can be stored in a variable and inspected:

```{r}
var <- ~toupper(letters[1:5])
var

# You can extract its expression:
rlang::f_rhs(var)

# Or inspect its enclosure:
rlang::f_env(var)
```

In the tidyeval framework, when a function takes an argument by
expression, it gets it in the form of a quosure. This way, it keeps
track of the context where the user typed the expression and knows
where to find user-defined variables mentioned in the quoted
expression. This is how dplyr is able to compute expressions that
refer to a mixture of data frame columns and user-defined variables:

```{r}
user_var <- 1000
mtcars %>% summarise(cyl = mean(cyl) * user_var)
```

However, things get more complicated if you want to store the name of
a column in a variable. While dplyr is able to understand variables
referring to _values_, it cannot understand variables referring to
_symbols_. It will just look up the content of the variable and get an
expression. In the following example, the function `mean()` gets the
quoted symbol `disp` as argument. Of course, `mean()` doesn't know
what to do with it since it expects a numeric vector:

```{r}
user_var <- ~disp
mtcars %>% summarise(user_var = mean(user_var) * 1000)
```

The solution to this problem is quasiquotation. This term refers to
the ability of unquoting part of an expression being captured.


## Unquoting

All expression-based functions in the tidyeval framework support
unquoting. Unquoting cancels quotation of parts of an expression. The
tidyverse provides unquoting operators in the functional form, `UQ()`,
and in syntactic form with the double bang operator `!!`. The base R
functions for quoting an expression do not understand unquoting, so we
are going to use the function `quosure()` to illustrate
quasiquotation. `quosure()` takes its argument by expression,
processes the unquoting operators, and returns a quosure.

```{r}
library("rlang")

# Here we capture `letters` as a symbol:
quosure(toupper(letters))

# Here we evaluate `letters` and embed its value in the captured expression:
quosure(toupper(!! letters))
```

In that example we have unquoted a value within `toupper()`. It is
also possible to unquote other quoted expressions. Unquoting such
symbolic objects provides a powerful a way of manipulating expressions
at capture time. From the point of view of the capturing function,
there is no difference between a symbol typed manually or a symbol
unquoted programmatically:

```{r}
var1 <- quote(letters[1:5])
var2 <- quote(letters[11:15])

quosure(toupper(!! var1))

quosure(toupper(!! var2))
```

The final touch is that you can safely unquote quosures instead of raw
expressions because tidyeval functions know how to evaluate
quosures. This lets you refer to variables in the current context:

```{r}
my_mutate <- function(x) {
  mtcars %>%
    select(cyl) %>%
    slice(1:5) %>%
    mutate(cyl2 = cyl + (!! x))
}

var <- 100
expr1 <- ~var * 2
expr2 <- ~var / 2

my_mutate(expr1)
my_mutate(expr2)
```

In a way, `quosure()` is tidyeval's most fundamental expression-based
function. It captures its expression and returns it without any
modification, post-processing, or evaluation. But all tidyeval
functions like `mutate()` behave like `quosure()` when passed an
expression, which ensures their full programmability.


### Unquote-splicing

The second important operation is unquote-splicing. Its functional
form is `UQS()` and the syntactic shortcut is `!!!`. Like unquoting,
it evaluates part of a captured expression, but if the unquoted object
is a vector, each element of the vector becomes an argument in the
surrounding function call:

```{r}
quosure(list(!!! letters[1:5]))
```

A very useful feature of unquote-splicing is that the vector names
become argument names:

```{r}
quosure(list(!!! list(foo = 1L, bar = quote(baz))))
```

This is of course handy to program with dplyr verbs taking named dots:

```{r}
args <- list(mean = ~mean(cyl * 20), count = ~n())
mtcars %>%
  group_by(am) %>%
  summarise(!!! args)
```


### Setting variable names

There are two main ways of programmatically setting the names of
arguments passed through dots. First you can splice a named vector as
mentioned above: You can also use the definition operator `:=` instead
of `=`. The former supports unquoting on its LHS. The unquoted operand
should evaluate to a string or a symbol:

```{r}
n <- 10
nms <- c("mean", "count")

summarise(mtcars,
  !! nms[[1]] := mean(mpg),
  !! nms[[2]] := n
)
```

Or equivalently with splicing:

```{r}
dots <- list(~mean(mpg), ~n)
named_dots <- rlang::set_names(dots, nms)

summarise(mtcars, !!! dots)

summarise(mtcars, !!! named_dots)
```
