---
title: "Programming with dplyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Programming with dplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(dplyr)

quo <- rlang::quosure
quos <- rlang::dots_quosures
enquo <- rlang::catch_quosure
quo_name <- rlang::quo_text
```

Most dplyr functions use non-standard evaluation (NSE). This is a catch-all term that means they don't follow the usual R rules of evaluation. Instead, they capture the expression that you typed and evaluate in a custom way. This has two main benefits for dplyr code:

*   Operations on data frames can be expressed succinctly because
    in expressions like `filter(df, x == 1, y == 2, z == 3)` dplyr
    assumes that `x`, `y`, and `z` are variables in df.
    
*   dplyr can choose to compute results in a different way to base R.
    For example, instead of doing computation itself, dplyr can generate
    the SQL code that will cause a remote database to do the work.

Unfortunately these benefits do not come for free. There are two main limitations:

*   Most dplyr arguments are not __referentially transparent__. That means
    you can't replace a value with an intermediate variable. In other words, this 
    code:
    
    ```{r, eval = FALSE}
    filter(df, my_var == 1)
    ```
    
    Is not equivalent to this code:
    
    ```{r, eval = FALSE}
    var <- my_var
    filter(df, var == 1)
    ```
    
    This makes it harder to create functions.
    
*   dplyr code is ambiguous. Depending on what variables are defined where, 
    `filter(df, x == y)` could be equivalent to any of:

    ```{r, eval = FALSE}
    df[df$x == df$y, ]
    df[df$x == y, ]
    df[x == df$y, ]
    df[x == y, ]
    ```
    
    This ambiguity makes functions that use dplyr more unpredictable
    than you might desire.

Fortunately, dplyr provides tools to overcome these challenges. They require a little more typing, but a small amount of upfront work is worth because they help you save time in the long run.

This vignette has two goals:

*   Show you show you how to use dplyr's __pronouns__ and __quasiquotation__
    to write reliable functions that reduce duplication in your data analysis 
    code.
    
*   To teach you about __quosures__ and the underyling theory of __tidyeval__.

## Programming recipes

The following recipes walk you through the basics of tidyeval, giving you some practical tools for handling duplication in dplyr code. The examples here are somewhat inauthentic because to make them easier to understand we've reduced them down to very simple components, to the point where writing a function almost isn't worth it. Your own data analysis code will be much more complicated, but if you look carefully you should be able to identify similar themes to the code below.

### Different data sets

If you want to repeat the same operation on different data frames, you can already do it:

```{r, eval = FALSE}
mutate(df1, y = a + x)
mutate(df2, y = a + x)
mutate(df3, y = a + x)
mutate(df4, y = a + x)
```

```{r}
mutate_y <- function(df) {
  mutate(df, y = a + x)
}
```

That's because the `.data` argument is referentially transparent. However, this simple approach has some drawbacks: it can fail silently if one of the variables isn't present in the data frame, but is present in the global environment.

```{r}
df1 <- tibble(x = 1:3)
a <- 10
mutate_y(df1)
```

You can overcome that problem by using the `.data` pronoun. This will (should) throw an error if the variable doesn't exist:

```{r, error = TRUE}
mutate_y <- function(df) {
  mutate(df, y = .data$a + .data$x)
}

mutate_y(df1)
```

This also prevents an `R CMD check` NOTE if you are including this code in a package (you'll also need to import `rlang::.data` to fully suppress the note).

### Different expressions

Working with the arguments is more challenging because they are not referentially transparent. Let's start with a simple case: you want to vary the name of the grouping variable.

```{r}
df <- tibble(
  g1 = c(1, 1, 2, 2, 2),
  g2 = c(1, 2, 1, 2, 1),
  a = sample(5), 
  b = sample(5)
)

df %>%
  group_by(g1) %>%
  summarise(a = mean(a))

df %>%
  group_by(g2) %>%
  summarise(a = mean(a))
```

We can't easily turn this into a function because this doesn't work:

```{r, error = TRUE}
var <- g1
```

Instead we need to capture, or __quote__ this expression. There are two ways to quote in base R: `quote()` and the formula, `~`.

```{r}
quote(g1)
~g1
```

For various reasons, neither of these does quite what we want, so instead we'll use `quo()` (this comes from the rlang package but is also exported by dplyr).

```{r}
var <- quo(g1)
var
```

This creates a __quosure__. A quosure is a special type of formula that you'll learn about later.

Now that we've captured this expression, how do we use it with `group_by()`? The naive approach doesn't work because `group_by()` expects its arguments to be names of existing variables:

```{r, error = TRUE}
df %>%
  group_by(var) %>%
  summarise(a = mean(a))
```

In other words, `group_by()` automatically quotes its inputs. But `var` is already quoted, so we need some way to __unquote__ it. In the tidyeval framework, you unquote an expression by putting `!!` in front of it.

```{r}
df %>%
  group_by(!!var) %>%
  summarise(a = mean(a))
```

We are now almost able to write the function:

```{r, error = TRUE}
my_summarise <- function(df, group_by) {
  group_by <- quo(group_by)
  print(group_by)
  
  df %>%
    group_by(!!group_by) %>%
    summarise(a = mean(a))
}

my_summarise(df, g1)
my_summarise(df, g2)
```

I've added a `print()` call to make it obvious what's going wrong here: `quo(group_by)` always returns `~group_by`. We want it to use the value that the user supplied so we need a variant `enquo()`:

```{r}
my_summarise <- function(df, group_by) {
  group_by <- enquo(group_by)

  df %>%
    group_by(!!group_by) %>%
    summarise(a = mean(a))
}

my_summarise(df, g1)
```

(If you're familiar with `quote()` and `substitute()` in base R, `quo()` is equivalent to `quote()` and `enquo()` is equivalent to `substitute()`.)

You might wonder how to extend this to handle multiple grouping variables: we'll come back to that a little later.

### Different input variable

Now let's tackle something a bit more complicated. The code below shows a duplicate `summarise()` statement where we compute three summaries, varying the input variable.

```{r}
summarise(df, mean = mean(a), sum = sum(a), n = n())
summarise(df, mean = mean(a * b), sum = sum(a * b), n = n())
```

To turn this into a function, we start in the same way: by quoting the variable with `quo()`, then unquoting it in the dplyr call with `!!`.

```{r}
var <- quo(a)
summarise(df, mean = mean(!!var), sum = sum(!!var), n = n())
```

You can also wrap `quo()` around the dplyr call to see what the generated call will look like.

```{r}
quo(summarise(df, 
  mean = mean(!!var),
  sum = sum(!!var),
  n = n()
))
```

This is the reason we use `quo()` instead of `~`: `quo()` also does unquoting. This makes it useful for previewing the expression the dplyr will execute.

Now we can turn our code into a function (remembering to replace `quo()` with `enquo()`), then check that it works:

```{r}
my_summarise <- function(df, expr) {
  expr <- enquo(expr)
  
  summarise(df, 
    mean = mean(!!expr),
    sum = sum(!!expr),
    n = n()
  )
}
my_summarise(df, a)
my_summarise(df, a * b)
```

### Different input and output variable

The next challenge is to vary the name of the output variables:

```{r}
mutate(df, mean_a = mean(a), sum_a = sum(a))
mutate(df, mean_b = mean(b), sum_b = sum(b))
```

This code is similar to the previous example, but there are two new wrinkles:

* We create the new names by pasting together strings, so
  we need `quo_name()` to convert the input expression to a string.

* `!!mean_name = mean(!!expr)` isn't valid R code, so we need to
  use the `:=` helper provided by rlang.

```{r}
my_mutate <- function(df, expr) {
  expr <- enquo(expr)
  mean_name <- paste0("mean_", quo_name(expr))
  sum_name <- paste0("sum_", quo_name(expr))
  
  mutate(df, 
    !!mean_name := mean(!!expr), 
    !!sum_name := sum(!!expr)
  )
}

my_mutate(df, a)
```

### Capturing multiple variables

It would be nice to extend our first example to accept any number of grouping variables. We need to make three changes:

*   Use `...` in the function argument so our function can accept any number
    of arguments.
    
*   Use `quos()` to capture all the `...` as a list of formulas.

*   Use `!!!` instead of `!!` to __splice__ the arguments into the 
    current call site.

```{r}
my_summarise <- function(df, ...) {
  group_by <- quos(...)

  df %>%
    group_by(!!!group_by) %>%
    summarise(a = mean(a))
}

my_summarise(df, g1, g2)
```

`!!!` takes a list of elements and splices them into to the current call:

```{r}
args <- list(na.rm = TRUE, trim = 0.25)
quo(mean(x, !!! args))

args <- list(quo(x), na.rm = TRUE, trim = 0.25)
quo(mean(!!! args))
```

Now that you've learned the basics of tidyeval through some practical examples, we'll dive into the theory. This will help you generalise learned here to new situations.

## Quoting

Quoting is the action of capturing an expression instead of evaluating it. All expression-based functions quote their arguments and get the R code as an expression rather than the result of evaluating that code. If you are an R user, you probably quote expressions on a regular basis. One of the most important quoting operators in R is the
_formula_. It is famously used for the specification of statistical models:

```{r}
disp ~ cyl + drat
```

The other quoting operator in base R is `quote()`. It returns a raw
expression rather than a formula:

```{r}
# Computing the value of the expression:
toupper(letters[1:5])

# Capturing the expression:
quote(toupper(letters[1:5]))
```

(Note that despite being called the double quote, `"` is not a quoting operator, because it generates a string, not an expression.)

In practice, the formula is the better of the two options because captures both the code, and its execution __environment__. This important because even simple expression can yield different values in different environments. For example, the `x` in the following two expressions refers to different values:

```{r}
f <- function(x) {
  ~ x
}

x1 <- f(10)
x2 <- f(100)
```

If you print out those formulas, it might look like they're the same. But look carefully at the environments --- they're different.

```{r}
x1
x2
```

When we evaluate those formulas using `rlang::eval_tidy()`, we see that they yield different values:

```{r, message = FALSE}
library(rlang)

eval_tidy(x1)
eval_tidy(x2)
```

This is an important property of R: a name can refer to different values in different environments.

When an object keep tracks of an environment, it is said to have an enclosure. This is the reason that functions in R are sometimes referred to as closures:

```{r}
typeof(mean)
```

For this reason we use a special name to refer to one-sided formulas: __quosures__. One-sided formulas are quotes (they carry an expression) that bundle their environment. Quosures are regular R objects. They can be stored in a variable and inspected:

```{r}
var <- ~toupper(letters[1:5])
var

# You can extract its expression:
f_rhs(var)

# Or inspect its enclosure:
f_env(var)
```

The tidyeval framework is characterised by its use of quosures. Using `enquo()` to capture a quosure is safer than the base R equivalent of using `substitute()` to capture only the expression. This is how dplyr is able to compute expressions that refer to a mixture of data frame columns and user-defined variables:

```{r}
user_var <- 1000
mtcars %>% summarise(cyl = mean(cyl) * user_var)
```

Automatic quoting makes dplyr very convenient for interactive use. But if you want to program with dplyr, you need some way to refer to variables indirectly. The solution to this problem is __quasiquotation__, which refers to the ability to unquoting part of an expression that is otherwise quoted. 

## Quasiquotation

> Put simply, quasi-quotation enables one to introduce symbols that stand for 
> a linguistic expression in a given instance and are used as that linguistic 
> expression in a different instance.
--- [Willard van Orman Quine](https://en.wikipedia.org/wiki/Quasi-quotation)

Quasiquotation was coined by Willard van Orman Quine in the 1940s, and was adopted for programming by the LISP community in the 1970s. All expression-based functions in the tidyeval framework support quasiquotation. Unquoting cancels quotation of parts of an expression. There are three types of unquoting:

* basic
* unquote splicing
* unquoting names

### Unquoting

The tidyverse provides unquoting operators in a functional form, `UQ()`, and as syntactic-sugar with `!!`.  As shown above, you can use `~` to create a quosure, but `~` does not do quasiquotation, so we'll need to use `quo()` to see how `UQ()` and `!!` work.

```{r}
# Here we capture `letters[1:5]` as an expression:
quo(toupper(letters[1:5]))

# Here we capture the value of `letters[1:5]`
quo(toupper(!!letters[1:5]))
```

It is also possible to unquote other quoted expressions. Unquoting such
symbolic objects provides a powerful a way of manipulating expressions. 

```{r}
var1 <- quo(letters[1:5])
quo(toupper(!!var1))
```

You can safely unquote quosures because they track their environments, and tidyeval functions know how to evaluate them. This allows any numbers of levels quoting and unquoting.

```{r}
my_mutate <- function(x) {
  mtcars %>%
    select(cyl) %>%
    slice(1:4) %>%
    mutate(cyl2 = cyl + (!! x))
}

f <- function(x) quo(x)
expr1 <- f(100)
expr2 <- f(10)

my_mutate(expr1)
my_mutate(expr2)
```

In a way, `quo()` is tidyeval's most fundamental expression-based
function. It captures its expression and returns it without any
modification, post-processing, or evaluation. But all tidyeval
functions like `mutate()` behave like `quosure()` when passed an
expression, which ensures their full programmability.

### Unquote-splicing

The second important operation is unquote-splicing. Its functional
form is `UQS()` and the syntactic shortcut is `!!!`. Like unquoting,
it evaluates part of a captured expression, but if the unquoted object
is a vector, each element of the vector becomes an argument in the
surrounding function call:

```{r}
quosure(list(!!! letters[1:5]))
```

A very useful feature of unquote-splicing is that the vector names
become argument names:

```{r}
quosure(list(!!! list(foo = 1L, bar = quote(baz))))
```

This is of course handy to program with dplyr verbs taking named dots:

```{r}
args <- list(mean = ~mean(cyl), count = ~n())
mtcars %>%
  group_by(am) %>%
  summarise(!!! args)
```

### Setting variable names

There are two main ways of programmatically setting the names of
arguments passed through dots. First you can splice a named vector as
mentioned above: You can also use the definition operator `:=` instead
of `=`. The former supports unquoting on its LHS. The unquoted operand
should evaluate to a string or a symbol:

```{r}
mean_nm <- "mean"
count_nm <- "count"

mtcars %>%
  group_by(am) %>%
  summarise(
    !!mean_nm := mean(cyl),
    !!count_nm := n()
  )
```
