---
title: "dplyr 0.8.0"
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(dplyr)
```

# More consistent use of factors

## Motivation

The work on making factors used more respectfully originates from the issue 341, 
which has been waiting for attention for the past four years. `#341` identified the 
need to take care of empty groups. 

Empty groups can arise from two situations:
 - when one of the grouping variable in `group_by()` is a factor and one of its levels has no data, e.g.
 
```{r}
tibble(
  x = 1:2, 
  f = factor(c("a", "b"), levels = c("a", "b", "c"))
) %>% 
  group_by(f)
```

The factor `f` has 3 levels, but only two are present in the data. 

 - when all of the data from one group is `filter()`ed out, e.g.
 
```{r}
tibble( x = 1:3, f = factor(c("a", "b", "c"))) %>% 
  group_by(f) %>% 
  filter(x < 2)
```

In that case, the grouped data before the filter has one row per level of `f`, and the filter 
only keeps the first row so makes 2 empty groups.  

## Previous behaviour
 
Older versions of `dplyr` did not make empty groups, because: 
- `group_by()` was building the grouping metadata only from the rows of the data, i.e. ignoring 
  the conceptual grouping structure. 
- `filter()` was making a lazily grouped tibble, recording only the names of grouping variables, 
  without producing the metadata, which was automatically made by a subsequent `group_by` whenever 
  this was necessary in the future. 
  
## Recusive slicing grouping algorithm

A new grouping algorithm, inspired from `tidyr::complete` is used in `dplyr` 0.8.0 to solve the first issue. 
The algorithm recursively goes through the grouping variables. When a grouping variable is a factor, 
the groups are made from its levels. On any other variable (`character`, `integer`, ...) the groups
are made from the unique values. 

Let's have a look at some examples, we'll use `tally()` to reveal the grouping structure and counts of groups:

```{r}
df <- tibble(
  x = c(1,2,1,2), 
  f = factor(c("a", "b", "a", "b"), levels = c("a", "b", "c"))
)
df %>% 
  group_by(f) %>% 
  tally()
```

In this first example, we group by a factor, so we get as many groups as the number of factors. 

```{r}
df %>% 
  group_by(f, x) %>% 
  tally()
```

Here we group by the factor `f` and the numeric vector `x`. Again we get 3 groups, because for the levels
"a" and "b" of `f`, there is only one value of `x`. The third group, associated with the level "c" 
sets the value of `x` to `NA` out of thin air. We call this a *sentinel NA* and we might make it obvious later that 
this is not the same as if we had a missing value in the data. 

```{r}
df %>% 
  group_by(x, f) %>% 
  tally()
```

In this case, we get more groups, and consequently more empty groups, because of the recursive slicing, 
first we find 2 unique values for the variable `x` (1 and 2), then we group by the factor `f` 
and therefore get 3 groups (because 3 levels) for each unique value of `x`. 

## Propagation of the grouping structure 

`filter()` has been reworked to respect the grouping stucture and gains the `.preserve` argument to control 
which groups to keep. 

When `.preserve` is set to `TRUE` (the default) the groups of the filtered tibble are the same as the groups 
of the original tibble. 

```{r}
df %>% 
  group_by(x, f) %>% 
  filter(x == 1) %>% 
  tally()

df %>% 
  group_by(f, x) %>% 
  filter(x == 1) %>% 
  tally()
```

When `.preserve` is set to `FALSE` the grouping structure is recalculated after the filtering. 

```{r}
df %>% 
  group_by(x, f) %>% 
  filter(x == 1, .preserve = FALSE) %>% 
  tally()
```

Here we only get 3 groups, from the 3 levels of `f` within the unique value of `x`

```{r}
df %>% 
  group_by(f, x) %>% 
  filter(x == 1, .preserve = FALSE) %>% 
  tally()
```

In that case, we get 3 groups, but the values of `x` are slightly different, i.e. the value of 
`x` associated with the level "b" in the empty group is a sentinel NA. 

# Tidy grouping structure

Previous versions of `dplyr` used a messy collection of attributes in the "grouped_df" class, which did not 
make it easy to reason about. `dplyr` 0.8.0 structures all the grouping information in a tibble with 
`n+1` columns (where `n` is the number of grouping variables) in the "groups" attribute. 

```{r}
df %>% 
  group_by(f, x) %>% 
  attr("groups")
```

The first columns identify the data for each of the group, one row per group. This is 
equivalent to the "labels" attribute used in previous versions of `dplyr`. 

The last column, always called `.rows` is a list column of integer vectors (possibly of length 0
for empty groups) identifying the indices of all the rows in the data that belong to the group. This 
is equivalent to the "indices" attribute used in previous versions. 

This grouping stucture tibble (maybe a gribble) can be retrieved by accessing the `groups` attribute, 
or preferably by using the `group_data()` generic, which has methods for ungrouped and row wise data too. 

```{r}
group_data(df)
group_data(group_by(df, f))
group_data(rowwise(df))
```

Similarly, the indices themselves can be retrieved using the `group_rows()` 

```{r}
group_rows(df)
group_rows(group_by(df, f))
group_rows(rowwise(df))
```

Having a consistent representation of the grouping structure makes it easier to reason about, and 
might open opportunities to discuss alternative ways of grouping

# Hybrid evaluation cleanup

**pending**

The initial goal for hybrid evaluation was to bypass potentially expensive R evaluation, and replace
it with more efficient C++ code. Unfortunately, there are situations where hybrid evaluation 
creates problems. 

There are two forms of hybrid evaluation in dplyr at the moment: full hybrid evaluation and 
hybrid folding. 

## Full hybrid evaluation

When the entire (summarise or mutate) expression can be handled, e.g. in `group_by(...) %>% summarise(m = mean(x))` 
the `mean` hybrid handler takes care of everything, i.e. calculate the mean of x for each group and 
structure that into a numeric vector. 

This does not need to allocate memory for each subset of `x` or the result of `mean(x)`. In addition, 
because it is dispatched internally, it does not need to pay the expensive price of S3 dispatch of the 
`mean` generic function from `base::`. 

This is where hybrid evaluation really makes a difference. Currently this is driven by a set of 
C++ classes inheriting from the virtual class `Result`, which is used for summary functions 
(such as `mean`) and window functions (such as `lead`). 

The proposal here is to rebase hybrid handlers on two virtual class (maybe templates) instead 
of one:
- `template <int RTYPE> Window<RTYPE>` would give a vector of type RTYPE of the right size. 
- `template <int RTYPE> Summary<RTYPE>` would summarise into of value of the right type. 

`mutate` and `summarise` would recognise expressions that are hybridable, and use the 
information to allocate the result then iterate through the groups to fill the result. 

This needs careful refactoring. We believe that this will make the code much simpler, 
with the consequence that it will be easier to write new hybrid handlers, i.e. we can imagine 
something like `x == 2` to be handled hybridly in `filter` by using a class deriving 
from `Window<LGLSXP>`. 

## Hybrid folding 

This is where hybrid evaluation creates problems, because it is sometimes too eager, 
and generally cannot faithfully mimic standard R evaulation. The original idea 
was to handle *parts*  of the expression using the hybrid handlers, e.g. in the 
expression `%>% group_by(...) %>% summarise(m = 1 + mean(x))` we would handle
`mean(x)` with the hybrid handler for `mean`, fold that into the expression and then 
fall back to r evaluation once we can no longer hybrid evaluate anything. 

Folding cannot be done once and for all groups, it is performed 
(including going through the expressions) for each group which has a price, we have to 
end with an R evaluation anyway, and then 
after that we still have no idea of what the result will be, so we collect and coerce 
the result with care. 

This has been the source of most of the "surprises" and also comes at a huge 
cost in terms of code complexity, and therefore maintainability. 

The proposal here is to totally abandon hybrid folding and replace it with an 
approach based on regular R evaluation. Expressions would be evaluated in an 
environment in which the names of the columns are mapped to their subsets
in the current group, and where functions such as `n()` and `row_number()` 
produce the desired result. 

Letting go of hybrid folding and making it easier to implement full hybrid handlers
will make hybrid evaluation simpler, more robust and less surprising. 

